---
title: "Project_(631)"
format: pdf
editor: visual
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Our research question is - **"How do Storage Capacity and RAM Size affect laptop price, and is there an interaction between them?"**


Loading the dataset and necessary libraries
```{r}

laptop <- read.table("Laptop_price.csv", header = TRUE, sep = ",")

library(dplyr)
library(ggplot2)
library(car)
```
Displaying summary statistics and first 6 rows of the dataset
```{r}
summary(laptop)
head(laptop)
```


Converting columns to factors for categorical analysis

```{r}
laptop <- laptop %>%
  mutate(across(c(Storage_Capacity, RAM_Size, Brand), as.factor))
```
Checking whether our dataset is balanced or not
```{r}
tapply(laptop[-1,]$Price,
laptop[-1,]$Storage_Capacity:laptop[-1,]$RAM_Size, length)
```
Checking structure and data types of the dataset
```{r}
str(laptop)
```




```{r}


# Summarize mean price by group
summary_data <- laptop %>%
  group_by(RAM_Size, Storage_Capacity) %>%
  summarise(mean_price = mean(Price, na.rm = TRUE), .groups = "drop")

# Interaction plot
ggplot(summary_data, aes(x = RAM_Size, y = mean_price, color = Storage_Capacity, group = Storage_Capacity)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Interaction Plot: RAM Size × Storage Capacity",
       x = "RAM Size",
       y = "Mean Price",
       color = "Storage Capacity")

```

Boxplot to visualize distribution of Price by Storage_Capacity and RAM_Size
```{r}
ggplot(laptop, aes(x = Storage_Capacity, y = Price, fill = RAM_Size)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Effect of Factors on Dependent Variable",
       x = "Storage Capacity",
       y = "Price",
       fill = "RAM Size")
```



Fitting Two-Way ANOVA: Storage_Capacity and RAM_Size
```{r}
# Fitting Two-Way ANOVA: Storage_Capacity and RAM_Size
model <- aov(Price ~ Storage_Capacity * RAM_Size, data = laptop)
Anova(model,type=2)
```

A **p-value \< 0.05** indicates a **statistically significant** effect.\
\
For both Storage_Capacity and RAM_Size have p value less than 0.05.

-   **Storage_Capacity** has a strong and significant impact on laptop price.

-   **RAM_Size** also has a strong and significant impact on laptop price.




Fitting updated Two-Way ANOVA: Storage_Capacity and RAM_Size
```{r}
# Fitting updated Two-Way ANOVA: Storage_Capacity and RAM_Size
redmodel <- aov(Price ~ Storage_Capacity + RAM_Size, data = laptop)
Anova(redmodel,type=2)
```
Shapiro-Wilk test for normality of residuals
```{r}
shapiro.test(residuals(redmodel))

```



Plotting to check for assumptions
```{r}
plot(redmodel)


```

-   **Residuals vs Fitted** - linearity and equal variance (homoscedasticity) assumptions are met

-   **Q-Q plot** - Normality assumption is satisfied.

-   **Scale-Location** - Homoscedasticity assumption satisfied.

-   **Residuals vs Leverage** - No concerning outliers.

-   **All assumptions are satisfied**.


Levene’s test to check equality of variances
```{r}

leveneTest(Price ~ Storage_Capacity * RAM_Size, data = laptop)

#Equal variances are satisfied
```

**Residuals are normal**

Q-Q plot for reduced model residuals
```{r}
qqPlot(resid(redmodel))
```


Tukey HSD test for pairwise comparisons among groups
```{r}
tukey <- TukeyHSD(redmodel)
tukey
```


